---
title: "Lab_12_Report"
format: pdf
editor: visual
author: German Vega
---

### 1. Background

The data for this hands-on session comes from a published RNA-seq experiment where airway smooth muscle cells were treated with dexamethasone, a synthetic glucocorticoid steroid with anti-inflammatory effects (Himes et al. 2014).

Glucocorticoids are used, for example, by people with asthma to reduce inflammation of the airways (Figure 1). The anti-inflammatory effects on airway smooth muscle (ASM) cells has been known for some time but the underlying molecular mechanisms are unclear.

### 2. Bioconductor setup

```{r}
library(BiocManager)
library(DESeq2)
```

### 3. Import countData and colData

```{r}
# Complete the missing code
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```
```{r}
head(metadata)
```


> Q1. How many genes are in this dataset? 

A: There are 38694 genes in the dataset

```{r}
nrow(counts)  # total number of genes
```


> Q2. How many ‘control’ cell lines do we have? 

A: There are 4 control lines in the dataset

```{r}
#getting the number of control lines in the datset
control_lines <- metadata[metadata$dex == "control", ]
nrow(control_lines) 
```

### 4. Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```
```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% dplyr::select(control$id)
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

> Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 

A: Replacing it with rowMeans(control.counts) is more robust because it automatically adapts to however many samples are present, making the code reusable across datasets of any size.

```{r}
control.mean <- rowSums(control.counts) / ncol(control.counts)
```


> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

A: 

```{r}
# Base approach
treated <- metadata[metadata[,"dex"]=="treated",]
treated.mean <- rowMeans(counts[ ,treated$id])
head(treated.mean)
```
```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

A:

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 

A: It would be geom_point()

```{r}
library(ggplot2)

ggplot(meancounts, aes(x = control.mean, y = treated.mean)) +
  geom_point(alpha = 0.3) +
  xlab("Control") +
  ylab("Treated")
```

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

A: The argument log allows us to do this 

```{r}
plot(meancounts[,1], meancounts[,2], 
     xlab="Control", ylab="Treated", 
     log="xy")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```
```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

A:arr.ind=TRUE tells which() to return the result as a matrix with both row and column indices instead of just a single position number, so for a 2D object like meancounts, you get back which gene (row) and which sample (column) contains a zero.
We then take the first column to extract only the row indices (genes), since we want to filter out genes, we don't care which sample the zero appeared in.
unique() is needed because a gene with zeros in both samples would show up twice in the row index list (once per zero found). unique() removes those duplicates so that gene only gets removed once.

```{r}
zero.vals <- which(meancounts[,1:2] == 0, arr.ind=TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
nrow(mycounts)
```
```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

A: There are 250 up rergulated genes in the database greater than 2 fc level

```{r}
paste("Up:", sum(up.ind))
```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

A: There are 367 down regulated genes reater than the 2 fc level

```{r}
paste("Down:", sum(down.ind))
```

> Q10. Do you trust these results? Why or why not?

A: No, I do not think that these eesuta can be trusted
The analysis is based purely on fold change, which only tells you the magnitude of difference, and not whether that difference is statistically meaningful. A gene can have a large fold change simply due to random noise or high variability across sample, without a p-value we can not analyze if the data is truly significant.

### 5. Setting up for DESeq

```{r}
library(DESeq2)
citation("DESeq2")
```
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```
### 6. Principal Component Analysis (PCA)

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```
```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```
```{r}
# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```
### 7. DESeq analysis

```{r}
dds <- DESeq(dds)
```

```{r}
results(dds)
```

```{r}
res <- results(dds)
res
```
```{r}
summary(res)
```
```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```
### 8. Adding annotation data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```
```{r}
columns(org.Hs.eg.db)
```
```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```
```{r}
head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.

A:

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```
```{r}
write.csv(res[ord,], "deseq_results.csv")
```

### 9. Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```
```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```
```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```
```{r}
library(EnhancedVolcano)
```
```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```
### 10. Pathway analysis

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

# Examine the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 2)
```
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```
```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```
```{r}
# Look at the first three down (less) pathways
head(keggres$less, 3)
```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```
```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa05310", kegg.native=FALSE)
```

> Q12. Can you do the same procedure as above to plot the pathview figures for the top 2 down-reguled pathways?

A: To plot the top 2 down-regulated pathways, we repeat the same pathview() procedure used for the first pathway, but substitute in the pathway IDs obtained from the top 2 rows of keggres$less. The first pathway is hsa05310 and the second is hsa05332, we pass foldchanges as the gene data in both calls so that genes are colored by their log2 fold change. Running pathview() once with default settings produces a KEGG-native PNG, and running it again with kegg.native=FALSE produces a PDF version of the same figure.

### OPTIONAL: Plotting counts for genes of interest

```{r}
i <- grep("CRISPLD2", res$symbol)
res[i,]
```

```{r}
rownames(res[i,])
```

```{r}
plotCounts(dds, gene="ENSG00000103196", intgroup="dex")
```

```{r}
# Return the data
d <- plotCounts(dds, gene="ENSG00000103196", intgroup="dex", returnData=TRUE)
head(d)
```

```{r}
boxplot(count ~ dex , data=d)
```

```{r}
library(ggplot2)
ggplot(d, aes(dex, count, fill=dex)) + 
  geom_boxplot() + 
  scale_y_log10() + 
  ggtitle("CRISPLD2")
```































